Longitudinal Principal Manifold Estimation

### 2024-02-06 paper/main.tex
overleaf updates

We propose a longitudinal extension to principal manifold estimation, with the goal of recovering smooth, longitudinally meaningful manifold estimates over time. The proposed approach uses a smoothing spline to smooth over the coefficients of the principal manifold embedding function estimated at each time point. This smoothing mitigates the effects of random disturbances to the manifold between time points. A novel data augmentation approach is used to allow the use of principal manifold estimation on self-intersecting manifolds. We use simulation studies with several classes of manifolds to demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods. These improvements persist when considering varying between-time-point noise levels and the types and magnitudes of systematic change between time points. We then apply the proposed method to estimating the surfaces of hippocampi of healthy individuals in the Alzheimer’s Disease Neuroimaging Initiative dataset.

@@

Ultimately, high levels of variability are present at the subcortical level, both between study visits for the same individual and between individuals, regardless of the segmentation approach used, and may be particularly influential when using automated segmentation methods. Mitigating the extent of this variability is a priority from a statistical perspective to improve the inferences that can be made from segmented image data. To this end, in this article, we propose a manifold learning-based method to develop smooth estimates of the surfaces of subcortical structures over time. Manifold learning describes a set of statistical approaches to modeling high-dimensional data that are assumed to lie along a low-dimensional manifold. In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space. For example, \cite{yueParameterizationWhiteMatter2016} seeks to compute a parametric representation of the corpus callosum. To obtain the low-dimensional parameterization of the substructures of interest, the proposed approach extends the principal manifold estimation approach introduced by \cite{mengPrincipalManifoldEstimation2021}.

Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}), and Laplacian eigenmaps (\cite{belkin2003laplacian}), are ill-equipped to reach meaningful estimates across multiple time points. These methods consider time as an additional dimension in the data, i.e. they include time in the dimension reduction process. This treatment of the time dimension causes it to be incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension. Here, we propose an alternative approach to estimating manifolds over time. Rather than include time as another dimension in a manifold learning method, we use a process that maintains the interpretability of the time dimension. Specifically, we use the method proposed in \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point. We then build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold. We demonstrate that the regularization involved in this process mitigates the effects of variability between study visits.

@@

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{}
\end{figure}

>>>>

Alzheimer's disease (AD) is a neurogenerative disorder affecting, among others, the structure of the brain. Longitudinal magnetic resonance imaging (MRI) data is used to model trajectories of change in brain regions of interest to identify areas more susceptible to atrophy. Most methods for extracting surfaces of brain regions are applied to individual scans from study participants independently. As a result, there is a wide variability of shape and volume estimates of brain regions of interest over time in longitudinal studies resulting in major implications for biomarker estimation and modeling, especially when used in therapeutic clinical trials. To address this problem, we propose a longitudinal principal manifold estimation method, with the goal of recovering smooth, longitudinally meaningful manifold estimates of shapes over time. The proposed approach uses a smoothing spline to smooth over the coefficients of the principal manifold embedding function estimated at each time point. This smoothing mitigates the effects of random disturbances to the manifold between time points. A novel data augmentation approach is used to allow the use of principal manifold estimation on self-intersecting manifolds. We use simulation studies with several classes of manifolds to demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods. These improvements persist when considering varying between-time-point noise levels and the types and magnitudes of systematic change between time points. We then apply the proposed method to estimating the surfaces of hippocampi of healthy individuals in the Alzheimer’s Disease Neuroimaging Initiative dataset.

@@

Ultimately, high levels of variability are present at the subcortical level, both between study visits for the same individual and between individuals, regardless of the segmentation approach used, and may be particularly influential when using automated segmentation methods. Mitigating the extent of this variability is a priority from a statistical perspective to improve the inferences that can be made from segmented image data. To this end, in this article, we propose a manifold learning-based method to develop smooth estimates of the surfaces of subcortical structures over time. Manifold learning describes a set of statistical approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis" --- \textit{high-dimensional data tend to lie in the vicinity of a low-dimensional manifold} (\cite{fefferman2016testing}). Manifold learning approaches have found applications in various scientific fields, e.g., medical signal processing (\cite{shen2022robust}) and robotics (\cite{gao2023k}). In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space. For example, \cite{yueParameterizationWhiteMatter2016} seeks to compute a parametric representation of the corpus callosum. To obtain the low-dimensional parameterization of the substructures of interest, the proposed approach extends the principal manifold estimation approach introduced by \cite{mengPrincipalManifoldEstimation2021}.

Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}; \cite{wu2018think}), and Laplacian eigenmaps (\cite{belkin2003laplacian}), are ill-equipped to reach meaningful estimates across multiple time points. These methods consider time as an additional dimension in the data, i.e. they include time in the dimension reduction process. This treatment of the time dimension causes it to be incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension. Here, we propose an alternative approach to estimating manifolds over time. Rather than include time as another dimension in a manifold learning method, we use a process that maintains the interpretability of the time dimension. Specifically, we use the method proposed in \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point. We then build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold. We demonstrate that the regularization involved in this process mitigates the effects of variability between study visits.

@@

The hippocampus is a region with a relatively irregular shape, which may account for the differences in performance between the LPME algorithm when applied to simulated data and when applied to the hippocampus data. To understand how the approach performs when used with a more regularly shaped brain region, we also applied the PME and LPME algorithms to the surface of the left thalamus of the same healthy individuals described previously. Estimates of the surface of the thalamus were obtained from the MRI images using the same preprocessing and segmentation steps detailed above for the hippocampus data.

Considering the same individual shown in Figures \ref{fig:adni_result} and \ref{fig:adni_cross_section}, results of fitting the PME and LPME algorithms to the surface of the left thalamus are shown in Figures \ref{fig:adni_lthal_result} and \ref{fig:adni_lthal_cross_section}. While the segmentation estimates of the thalamus do not show the same sort of deviations that are clearly visible for the hippocampus, the cross section plots shown in Figure \ref{fig:adni_lthal_cross_section} indicate that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus.

@@

\begin{figure}%
  \centering
  \subfloat[\centering Data]{{\includegraphics[height=22cm]{adni_plots/adni_lthal_data_plot}}}%
  \hfill
  \subfloat[\centering LPME Estimate]{{\includegraphics[height=22cm]{adni_plots/adni_lthal_lpme_isomap_plot}}}
  \hfill
  \subfloat[\centering PME Estimate]{{\includegraphics[height=22cm]{adni_plots/adni_lthal_pme_plot}}}

  \caption{Left Thalamus, Cognitive Healthy ADNI Participant}
  \label{fig:adni_lthal_result}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{adni_plots/adni_lthal_cross_section.png}
    \caption{Left Thalamus Cross Section}
    \label{fig:adni_lthal_cross_section}
\end{figure}
